---
title: "Learning R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(RCurl)
library(ggplot2)

knitr::opts_chunk$set(echo = FALSE)

dos <- 2
pie <- 3.14
apple <- 12
calculate <- function(x) {
  y <- (x + 15) * 11 / 8
  return(y)
}
thanksgiving_pies <- read.csv(text=getURL("https://raw.githubusercontent.com/aekendig/CodingEnvSci_UF_FieldTrip/master/coding_curriculum/data/thanksgiving-pies-2015.csv"))
total <- sum(thanksgiving_pies$count)
```

```{r pie-percent, include=FALSE}
thanksgiving_pies$percent <- thanksgiving_pies$count/total*100
```

## Welcome to R!
R is a free software that scientists and other researchers use for understanding data that they collect. They use it for mathematical analyses, like statistics, and visualizations, like charts or graphs (we'll call them plots from now on). It is "open-source", which means that you can see what's "under the hood" and people across the world are able to contribute to it to make it better. 

The program that we're working in is called R Studio. This is a special version of R in which we can see all of our tools in front of us. 

* The top left box is the script - this is where we write our code. 
* The bottom left box is the console - this is where R reads our code and gives us output. 
* The top right box has a list of all of the "objects" that we create in R (we'll come back to that later)
* The bottom right box shows us our files, plots, packages, the help menu, and the Viewer tab to look at internet content.

For this first lesson, we'll work in this special interactive window. After this, you'll work directly in RStudio.

Let's get started!


## R is a calculator (and more)
At the most basic level, R operates just a like a calculator that you would use in math class. Try entering a mathematical expression below, such as 2+2 and press "Run Code"

```{r 2+2, exercise=TRUE}

```

But R is a pretty smart calculator. For example, we can assign a value to a letter or word and R will remember it for us. Use the code that we've provided below to assign the value 2 to the Spanish word "dos". When you run the code, there won't be any output.

```{r dos, exercise=TRUE, exercise.eval=TRUE}
dos <- 2
```

Now, we'll ask R what dos is. In the box, type dos and run the code (R may even offer to help complete your answer as you're typing):

```{r dos recall, exercise=TRUE}

```

Notice that R is case-sensitive, that is, a lowercase letter is different than an uppercase letter. Try pressing "Start Over" and type in Dos with an uppercase "D". What happens?

We can use the word "dos" whenever we want to use the number 2. Use R to figure out what dos + dos is. The output should be 4:

```{r using dos, exercise=TRUE}

```

You can use any combinations of words or letters as long as there are no spaces in between. For example, this works:

```{r variable_combo, exercise=TRUE}
I_love_horses <- 55
I_love_horses
```

This doesn't:

```{r variable combo, exercise=TRUE}
I love horses <- 55
I love horses
```

## Hashtag #ILoveRStudio
If you look back at your RStudio window, you'll notice that there are a lot of hashtags or pound signs (this symbol: #). Hashtags tell R: "Please don't read this as code" and can be useful for explaining to someone else what your code does. 

```{r hashtag, exercise=TRUE}
# save the number 3.14 as the word pie
pie <- 3.14
```

We call this a "comment".
Try writing your own comment for the code section below.

```{r comment, exercise=TRUE}

apple <- 12
```

What's apple x pie (use * for multiplication)?

```{r apple pie, exercise=TRUE}

```

## Functions
We usually want to do more complicated things with R than arithmetic. Functions make this easier. Functions take an input, change it, and provide an output. Let's work through an example.

Say that you want a function that takes a number, adds 15, multiplies by 11, and divides by 8. We'll name the function "calculate". This is how we would write that function:

```{r function, exercise=TRUE}
calculate <- function(x) {
  y <- (x + 15) * 11 / 8
  return(y)
}
```

To run the function, we need to give it an input (the value for x). Let's try it with 10:

```{r function input, exercise=TRUE}
calculate(10)
```

Now you try it for a number:

```{r function input 2, exercise=TRUE}

```

## Packages 
Packages are groups of functions that work well together, like a toolbox with many tools inside. We use the function "library" to load packages. Let's load the package "ggplot2", which is good for making plots, by running the code:

```{r packages, exercise=TRUE}
library(ggplot2)
```

Sometimes you'll see an output, such as information about the package, sometimes not. You can usually ignore it.

## It's data time
Now, let's practice working with some data. 

Often, we type data into a program like Excel. We have to then import it into R to work with it. Run the code below, which imports data on the types of pies people eat at Thanksgiving. The file is stored on a website, to we're using a url to get it.

```{r import data, exercise=TRUE}
thanksgiving_pies <- read.csv(text=getURL("https://raw.githubusercontent.com/aekendig/CodingEnvSci_UF_FieldTrip/master/coding_curriculum/data/thanksgiving-pies-2015.csv"))
```

We call data tables like this "dataframes". Let's look at the top of our dataframe:

```{r head, exercise=TRUE}
head(thanksgiving_pies)
```

The dataframe has two columns: pie_type and count (the number of people who chose that pie as their preferred type). In order to look at one column, we use a dollar sign ($). Let's look at the column of pie types:

```{r columns, exercise=TRUE}
thanksgiving_pies$pie_type
```

Now it's your turn. Tell R to print the column for "count" using the dollar sign:

```{r columns 2, exercise=TRUE}

```

We can use R's "calculator capabilities" to manipulate the dataframe. For example, let's say we want the percentage of people rather than the count. First, let's calculate the total number of people who answered the survey by adding up all of the counts and assigning the value to the word "total":

```{r total, exercise=TRUE}
total <- sum(thanksgiving_pies$count)
total
```

Next, we divide each count by the total and multiply by 100 to get the percent. We'll save this as a new column called "percent":

```{r percent, exercise=TRUE}
thanksgiving_pies$percent <- thanksgiving_pies$count/total*100
thanksgiving_pies$percent
```

Now, let's look at the percentage of people who prefer each pie type by making a plot:

```{r plot, exercise=TRUE, exercise.setup="pie-percent"}
ggplot(thanksgiving_pies, aes(x = pie_type, y = percent)) +
  geom_bar(stat = "identity") 
```

Let's walk through the different pieces of that code. First, we use the function ggplot to make our plot. We tell R that the dataframe is thanksgiving_pies. Then, we use different parts of that dataframe to make the plot: pie_type is on the x-axis and percent is on the y-axis. Then, we tell R that we want to make a bar chart (geom_bar(stat = "identity"), the stat = "identity" part is specific to bar charts).

Try making a similar plot with "count" on the y-axis. You are encouraged to copy, paste, and edit code:

```{r plot 2, exercise=TRUE, exercise.setup="pie-percent"}

```